---
title: "Stratigaphic diagrams"
subtitle: "PalaeOpen Training School — Sofia, 15–18 June 2026"
author: "Xavier Benito"
execute: 
  echo: true
  warning: false
  message: false
  error: false
format:
  html:
    toc: true #table of contents
    toc-location: right
    date-format: long
    date-modified: last-modified
    code-copy: true
    code-fold: show
    code-line-numbers: true
    theme: flatly
    embed-resources: true #to allow portability of html files
---


## Learning Objectives

1.  
7.  Build reproducible workflows using Quarto

## Online resources

1.  [Code repository for the workshop](https://github.com/PalaeOpen/workshop-neotoma2-stratplots-trainingschool)
2.  [Neotoma Palaeoecology Database Manual](https://open.neotomadb.org/manual/)
3.  [neotoma2 GitHub Repository](https://github.com/neotomadb/neotoma2 (including previous Neotoma workshop repositories)
4.  [PalaeOpen R Zulip channel](https://palaeopen.zulipchat.com/#narrow/channel/540741-it_r)
5.  [Neotoma Slack channel](https://neotomadb.slack.com/join/shared_invite/zt-cvsv53ep-wjGeCTkq7IhP6eUNA9NxYQ#/shared-invite/email) 

## Install and load the necessary packages

This workflow relies on different R libraries, including tidyverse, geojsonsf and leaflet, to handle geographic data and interactive mapping. To streamline the setup, we utilize the pacman management tool. If any required package is missing from your local environment, pacman will automatically download and install it before loading.

```{r}
# Clean workspace first
rm(list = ls())
#install.packages("pacman") #install the package to load the necessary packages
pacman::p_load(neotoma2, tidyverse, tidypaleo, ggplot2, sf, geojsonsf, leaflet, raster, egg, openxlsx, cowplot)
```

# Stratigraphic plot
There exists several packages to perform stratigraphic diagrams of your data. The data needs to be on long or wide format as a function of the package used:

| Package | Data structure | function |
| --------- | ----------- | ----------- |
| *ggplot2* | long format | `ggplot()`|
| *tidypaleo* | long format | `Stratiplot()`|
| *analogue* | wide format | `ggplot()`|
| *rioja* | wide format | `strat.plot()`|
| *riojaPlot* | wide format | `riojaPlot()`|

In this tutorial we will use two different packages: [`tidypaleo`](https://paleolimbot.github.io/tidypaleo/index.html) and [`riojaPlot`](https://nsj3.github.io/riojaPlot_gallery/riojaPlot_Gallery.html#1)

`tidypaleo` can be installed directly from CRAN (`install.packages("tidypaleo)`) and `riojaPlot` is not yet on CRAN. To install it, you can run:

```{r install riojaPlot}
remotes::install_github("nsj3/riojaPlot", built_vignettes=TRUE, dependencies=TRUE)
# or
install.packages("riojaPlot", repos="https://nsj3.r-universe.dev")
```

As we did in the [intro to neotoma2 workshop](https://github.com/PalaeOpen/workshop-neotoma2-trainingschool), we will use the site **Prášilské jezero**. This site have available datasets for **Pollen**, **diatoms**, **chironomids** & **charcoal**.

We first obtain the site metadata including all its datasets. Then we filter for datasets **Pollen**, **diatoms** and **chironomids**
```{r}
pra_site <- get_datasets(get_sites(sitename = "Prášilské jezero"), all_data = TRUE) %>%
  neotoma2::filter(datasettype %in% c("pollen","diatom","chironomid"))
```

And we extract the sample data:
```{r}
pra_rec <- pra_site %>% get_downloads()
datasets(pra_rec)
pra_samp <- pra_rec %>% samples()

```

Let's say we are interested in creating an stratigraphic diagram with the aquatic and terrestrial datasets of the site. 

# Option 1: `tidypaleo`. This package is basically built in `ggplot2` so we need a long-format dataset

## Diatoms
```{r}
pra_samp_diat <- samples(pra_rec) %>% 
  dplyr::filter(ecologicalgroup=="DIAT") %>%
  dplyr::filter(taxongroup == "Diatoms")
```

Calculate diatom proportions:
```{r}
pra_diat_perc <- pra_samp_diat %>%
  dplyr::group_by(depth, age) %>%
  dplyr::mutate(diatcount = sum(value, na.rm = TRUE)) %>%
  dplyr::group_by(variablename) %>% 
  dplyr::mutate(relative_abundance_percent = value / diatcount) %>% 
  dplyr::arrange(desc(age))
```

Prepare the dataset and plot
```{r}
diatoms_long <- pra_diat_perc %>%
  dplyr::select(depth, age, variablename, relative_abundance_percent) %>% #select the variables of interest from the sample object
  dplyr::mutate(relative_abundance_percent = as.numeric(relative_abundance_percent)) %>%
  rename(taxa=variablename) %>% #rename column variablename to taxa
  mutate(taxa = factor(taxa)) 

levels(diatoms$taxa)

diat_coniss <- diatoms_long %>%
  as.tibble() %>%
  nested_data(qualifiers = age, key = taxa, value = relative_abundance_percent) %>%
  nested_chclust_coniss()

# Here we select the top 15 most abundant diatom taxa
site_plot_diat <- taxa(pra_rec) %>% dplyr::filter(taxongroup == "Diatoms") %>%
  dplyr::arrange(desc(samples)) %>% 
  head(n = 15)

theme_set(theme_bw(base_size=12))
diat_plot <- diatoms_long %>%
  filter(taxa %in% site_plot_diat$variablename) %>%
  ggplot(., aes(x = relative_abundance_percent*100, y = age)) +
  geom_col_segsh() +
  scale_y_reverse() +
  facet_abundanceh(vars(taxa)) +
  labs(y = "Age (cal yr BP)") 
diat_plot

diat_coniss_plt <- ggplot() +
  layer_dendrogram(diat_coniss, aes(y = age)) +
  scale_y_reverse() +
  facet_geochem_gridh(vars("CONISS")) +
  labs(x = NULL)

patchwork::wrap_plots(
  diat_plot + 
    theme(strip.background = element_blank(), strip.text.y = element_blank()),
  diat_coniss_plt +
    theme(axis.text.y.left = element_blank(), axis.ticks.y.left = element_blank()) +
    labs(y = NULL),
  nrow = 1,
  widths = c(6, 1)
)
```

## Pollen
```{r}
pra_samp_pollen <- samples(pra_rec) %>% 
  dplyr::filter(ecologicalgroup==c("TRSH", "UPHE", "VACR")) %>%
  dplyr::filter(elementtype %in% c("pollen", "spore")) %>%
  dplyr::filter(units == "NISP") 
```

Calculate pollen proportions:
```{r}
pra_pollen_perc <- pra_samp_pollen %>%
  dplyr::group_by(depth, age) %>%
  dplyr::mutate(pollencount = sum(value, na.rm = TRUE)) %>%
  dplyr::group_by(variablename) %>% 
  dplyr::mutate(relative_abundance_percent = value / pollencount) %>% 
  dplyr::arrange(desc(age))
```

Prepare the dataset and plot
```{r}
pollen <- pra_pollen_perc %>%
  dplyr::select(depth, age, variablename, relative_abundance_percent, ecologicalgroup) %>% #select the variables of interest from the sample object
  dplyr::mutate(relative_abundance_percent = as.numeric(relative_abundance_percent)) %>%
  rename(taxa=variablename) %>% #rename column variablename to taxa
  mutate(taxa = factor(taxa)) 

unique(pollen$taxa)
unique(pollen$ecologicalgroup)

# Here we select the top 20 most abundant pollen taxa
site_plot_pollen <- taxa(pra_rec) %>% 
  dplyr::filter(elementtype %in% c("pollen", "spore")) %>%
  dplyr::filter(units == "NISP") %>%
  dplyr::arrange(desc(samples)) %>% 
  head(n = 20)

theme_set(theme_bw(base_size=12))
pollen_plot <- pollen %>%
  filter(taxa %in% site_plot_pollen$variablename) %>%
  ggplot(., aes(x = relative_abundance_percent*100, y = age, color=ecologicalgroup)) +
  geom_col_segsh() +
  scale_y_reverse() +
  facet_abundanceh(vars(taxa)) +
  labs(y = "Age (cal yr BP)") 
pollen_plot
```

Create a composite plot for diatoms and pollen records
```{r}
plot_grid(diat_plot, pollen_plot,
                            nrow = 1, 
                            labels = c("Diatoms", "Pollen"), label_size = 12,
                            rel_widths = c(0.5,0.5), align = "hv") 
```

# Exercise: add the chironomid diagram next to the diatom stratigraphic plot

## Option 2: `riojaPlot`
**riojaPlot** package needs a wide-format dataset, so we first take the **diatom** long format table and convert it into a wide table:

```{r}
pra_diat_wide <- tidyr::pivot_wider(pra_diat_perc,
                                      id_cols = c(depth, age),
                                      names_from = variablename,
                                      values_from = relative_abundance_percent,
                                      values_fill = 0)
```

We plot the diatom stratigraphic diagram:
```{r}
riojaPlot(pra_diat_wide[,-1:-2]*100, pra_diat_wide[,1:2],
                       selVars = site_plot_diat$variablename,
                       scale.percent = TRUE,
                       sec.yvar.name="age",
                       plot.sec.axis = TRUE,
                       srt.xlabel = 60,
                       xRight = 0.85) 
```

# Adding multiple proxies with `riojaPlot`

calculate principal curves of pollen data and add it to the diatoms





